### 2025-06-04 writing/manuscript.tex
draft 3 intro lit rev

\pagebreak
\section{Introduction}
Equally qualified individuals may face very different labor market outcomes depending on their socioeconomic status \citep{stansbury2024class}. A key driver of this inequality is due to differences in social capital,\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} with recent empirical work characterizing its most important facet as the “share of high-SES friends among individuals with low-SES” as it correlates strongly with labor market income \citep{chetty2022na}. A lack of social capital means a lack of access to individuals with influential (higher paid) jobs and job opportunities. In economic terms, it implies having worse outcomes when using one's network to find jobs conditional on the capacity on leveraging one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use. \textcolor{red}{fill in citaiton sociology reading from slides}}

@@

Yet, experimental evidence shows referrals can be biased under substantial pay-for-performance incentives beyond what is attribuable to differences in network compositions, at least for the case of gender \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES may further exacerbate outcomes of low-SES individuals: If job information are in the hands of a select few high-SES which low-SES have already limited network access to (social capital hypothesis), and high-SES referrers are biased against low-SES, referring other high-SES at higher rates than their network composition, we should expect referral hiring to further disadvantagee low-SES. The empirical question we answer is whether there is a bias against low-SES once we account for the network SES compositon in a controlled setting.

@@

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. As our participants are students at the university and refer based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without the need for the referred candidate to take any action. This eliminates reputational concerns as there is no firm (see for example \citet{witte2021workers,bandiera2009social}). At the same time, national university entry exam scores are still objective, widely accepted measures of ability, and we show evidence that referrers in our setting possess accurate information about these signals.

In a university setting, class attendance provides essential opportunities for face-to-face interaction between students. On the one hand, this reduces network segregation by providing ample opportunities to meet across-SES, because of the exposure to an equal or higher level of high-SES compared to the population \citep{chetty2022n}. On the other hand, as students take more and more classes together, their similarities across all observable characteristics tend to increase \citep{kossinets_origins_2009}, which should drive the high- and low-SES networks to segregate. Our setting is ideal to study these opposing forces: First, The very high level of income inequality and existence of deeply rooted historical groups in Colombia makes SES differences extremely visible in access to tertiary education, where the rich and poor typically select into different institutions \citep{jaramillo-echeverri2023}. Yet, thanks to the particular standing of the institution we have chosen for this study (Figure \ref{school_quality_income}), different SES groups including both low- and high-SES mix together in this university.

@@

Do segregated networks account for all the differences in SES referral rates across SES groups? Same-SES referrals are 17\% more common than is suggested by referrer networks. High- and low-SES referrers under refer each other by 50\%. Yet, controlling for networks, our results suggest only low-SES students show significant same-SES homophily in referrals. Regardless of SES, referrals identify highly able individuals, and are characterized by a very high number of courses taken together. The latter underlies the impact of program selection in referrals, where the smaller and more homogenous choice sets come into play.
Our treatment randomized participants across two different incentive schemes by adding a substantial monetary bonus (\$25) for the referred candidate on top of the pay-for-performance incentives. We provide evidence that treatment incentives did not change the referral behavior across the same-SES referral rate, the number of courses taken together with the referral candidate, and the candidate's exam scores.

This paper extends various the literature

>>>>

\pagebreak
\section{Introduction}
Equally qualified individuals may face very different labor market outcomes depending on their socioeconomic status \citep{stansbury2024class}. A key driver of this inequality is due to differences in social capital,\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} with recent empirical work characterizing its most important facet as the “share of high-SES friends among individuals with low-SES” as it correlates strongly with labor market income \citep{chetty2022na}. A lack of social capital means a lack of access to individuals with influential (higher paid) jobs and job opportunities. In economic terms, it implies having worse outcomes when using one's network to find jobs conditional on the capacity on leveraging one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

@@

Yet, experimental evidence shows referrals can be biased even under substantial pay-for-performance incentives beyond what is attribuable to differences in network compositions, at least for the case of gender \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES may further exacerbate outcomes of low-SES individuals: If job information are in the hands of a select few high-SES which low-SES have already limited network access to (social capital hypothesis), and high-SES referrers are biased against low-SES, referring other high-SES at higher rates than their network composition, we should expect referral hiring to further disadvantagee low-SES. The empirical question we answer is whether there is a bias against low-SES once we account for the network SES compositon in a controlled setting.

@@

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. As our participants are students at the university and refer based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without the need for the referred candidate to take any action and without revealing the identity of the referrer. This eliminates reputational concerns as there is no hiring firm, and puts a lower bound on the expected reciprocity for the referrer in combination with pay-for-performance incentives \citep{witte2021workers,bandiera2009social}. At the same time, referring based on university entry exam scores are still an objective, widely accepted measure of ability, and we show evidence that referrers in our setting not only possess accurate information about these signals but are also able to screen more productive individuals from their university network.

In a university setting, class attendance provides essential opportunities for face-to-face interaction between students. On the one hand, this reduces network segregation by providing ample opportunities to meet across-SES, because of the exposure to an equal or higher level of high-SES compared to the population \citep{chetty2022n}.\footnote{In a different sample from the same university population, \citet{diaz_peer_2025} show this holds true for the highest-SES individuals at this institution, account for about 6\% of their sample but less than 5\% of Colombian high-school graduates \citet{fergusson2021desigualdad}.} On the other hand, as students take more and more classes together, their similarities across all observable characteristics tend to increase \citep{kossinets_origins_2009}, which should drive the high- and low-SES networks to segregate. Our setting is ideal to study these opposing forces: First, The very high level of income inequality and existence of deeply rooted historical groups in Colombia makes SES differences extremely visible in access to tertiary education, where the rich and poor typically select into different institutions \citep{jaramillo-echeverri2023}. Yet, thanks to the particular standing of the institution we have chosen for this study (Figure \ref{school_quality_income}), all SES groups including both low- and high-SES mix together in this university. Second, using administrative data, we are able to reconstruct 734 participants' complete university network based on the number of common courses they have taken together with other students. This allows directly identifying the individual characteristics of those getting referrals among all possible candidates, as well as descriptive characterizations of similarity (e.g., in same-SES share) in student networks as a function of the number of classes taken.

@@

Do segregated networks account for all the differences in SES referral rates across SES groups? Although same-SES referrals are 17\% more common than is suggested by referrer networks, controlling for these, we find no general SES-bias against beyond what is attribuable to network composition. Regardless of SES, participants refer productive individuals, and referred candidates are characterized by a very high number of courses taken together. The latter underlies the impact of program selection, where smaller and more homogenous parts of the networks are activated for referrals made in our setting. Our treatment randomized participants across two different incentive schemes by adding a substantial monetary bonus (\$25) for the referred candidate on top of the pay-for-performance incentives. We provide evidence that treatment incentives did not change the referral behavior across the same-SES referral rate, the number of courses taken together with the referral candidate, and the candidate's exam scores.

This paper contributes to the literature on referral experiments by solving the challenge of observing the entire referral network. Earlier research could only make observe referrals made across different incentive structures or experimental instructions and make according conclusions. For example, when participants are paid on the basis of their referred candidate's productivity instead of receiving a fixed finder's fee \citep{beaman_who_2012}, or when participants are restrictred to refer either a male or female candidate instead of freely \citep{beaman_job_2018}. \citet{pallais_why_2016} recruited a random sample of nonreferred workers to compare with referred ones, but none of the previous studies could provide a direct comparison of the referral choice set with those who were selected by participants. Closest to our work is the work of \citet{hederos2025gender}, who elicited friendship networks by asking referrers to name 5 friends. Their findings suggest only half of those who were referred were from the elicited friendship network, and thus is not a complete observation of the referral choice set. Although commonplace, censored elicitation methods also result in underestimating network effects \citep{griffith2022jle} and may suffer from biases in recall. We are able to take our analysis one step further by asking for referrals from the enrollment network, where we have complete information on every single connection that may or may not get a referral. This allows us to neatly separate the effect of the network composition from any potential biases stemming from the referral procedure itself.

 Second, we contribute to the earlier work on SES-biases in referrals. To our knowledge, \citet{diaz_peer_2025} are the first to study SES-biases in referals. Drawing from a similar sample from the same institution, they focus on referrals from first year students made within mixed-program classrooms, and find no evidence for an aggregate bias against low-SES. We also find no aggregate bias against low-SES in referrals. Our setup differs as we sample from students who completed their first year and impose no limits on referring from a classroom. This has several implications: We find that referrals in our setup go to individuals within the same program, and that programs have different SES-shares which become more even more accentuated as students take more courses together. While networks in our sample drive inequality in referral outcomes because of the institional environment, we have no reason to believe networks in \citet{diaz_peer_2025} have similar levels of segregation to begin with.

Finally, we contribute to the growing literature on SES differences in the labor market, expliciting the role of networks as a driver of inequality.
\citet{stansbury2024class} find that low-SES researchers coauthor more often with other low-SES, and have networks that have lower values which can explain why
